---
title: "Population Growth + Prosperity"
author: "Hunter Hall, Rebecca Elliott, Gio Rocco, Eric Jones, & Soumaila Ouedraogo"
date: "11/16/2019"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```
# **Introduction and Discussion of Ohio's Sources of Population Growth**
# We use Legatum's Prosperity Index and the US Census' population estimates for 2010 through 2018.

Legatum's Prosperity Index took a load of economic indicators from each state and compiled them into an index that allowed them to rank each state by their new prosperity indicator.

## **Examination of the aspects of Ohio's population growth***
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(reshape2) # for melting from wide to long dataframes
library(scales) # for setting graph scales
```

```{r}

# Loading the data 
# There is a Unicode UTF-8 BOM
# http://en.wikipedia.org/wiki/Byte_order_mark at the beginning
# of the file, which adds ï.. to the beginning of the first value. 
# Appending fileEncoding = "UTF-8-BOM" makes R ignore the file encoding
# and removes the additional characters.
setwd('C:/Users/Hunter Hall/Google Drive/Cincinnati/Class/eda_project/')
census_data <- read.csv("census data_eric.csv", fileEncoding = "UTF-8-BOM")
census<-as.data.frame(read.csv("census data.csv"))
prosperity<-as.data.frame(read.csv("prosperity.csv"))
pillars<-data.frame(read.csv("pillars.csv"))
state <- read.csv("Pillars_x11.csv")
state.data <- read.csv("census data.csv")
PI <- read.csv("prosperity index.csv")

```

```{r}
# lowering cases!

census_data$NAME <- tolower(census_data$NAME)
names(census_data) <- tolower(names(census_data))
```

```{r}
# Now we pull only out ohio data. 
# The 2010 entries are significantly different from the rest
# of the entries, so you can't compare them with the rest of the data.

ohio_data_raw <- census_data %>%
  select(-name, -contains("2010")) %>%
  slice(41)
```

```{r}
# Reformatting the data into a usable shape

titles <- c("year",
            "popestimate",
            "npopchg",
            "births",
            "deaths",
            "naturalinc",
            "internationalmig",
            "domesticmig",
            "netmig",
            "residual",
            "rbirth",
            "rdeath",
            "rnaturalinc",
            "rinternationalmig",
            "rdomesticmig",
            "rnetmig"
            )
y2011 <- ohio_data_raw %>%
  select(ends_with("2011"))
y2011 <- cbind("2011", y2011)
colnames(y2011) <- titles

y2012 <- ohio_data_raw %>%
  select(ends_with("2012"))
y2012 <- cbind("2012", y2012)
colnames(y2012) <- titles

y2013 <- ohio_data_raw %>%
  select(ends_with("2013"))
y2013 <- cbind("2013", y2013)
colnames(y2013) <- titles

y2014 <- ohio_data_raw %>%
  select(ends_with("2014"))
y2014 <- cbind("2014", y2014)
colnames(y2014) <- titles

y2015 <- ohio_data_raw %>%
  select(ends_with("2015"))
y2015 <- cbind("2015", y2015)
colnames(y2015) <- titles

y2016 <- ohio_data_raw %>%
  select(ends_with("2016"))
y2016 <- cbind("2016", y2016)
colnames(y2016) <- titles

y2017 <- ohio_data_raw %>%
  select(ends_with("2017"))
y2017 <- cbind("2017", y2017)
colnames(y2017) <- titles

y2018 <- ohio_data_raw %>%
  select(ends_with("2018"))
y2018 <- cbind("2018", y2018)
colnames(y2018) <- titles

ohio_data_yr <- rbind(y2011,
                      y2012,
                      y2013,
                      y2014,
                      y2015,
                      y2016,
                      y2017,
                      y2018
                      )

ohio_data_yr$year <- as.numeric(as.character(ohio_data_yr$year))
```


```{r}
# cleaning the environment

rm(ohio_data_raw,
   y2011,
   y2012,
   y2013,
   y2014,
   y2015,
   y2016,
   y2017,
   y2018,
   titles)
```

```{r}
# And now we plot our data

ohio_totals <- ohio_data_yr %>%
  select(year, npopchg, naturalinc:netmig) %>%
  melt(id.var = "year") %>%
  mutate(variable = replace(as.character(variable), variable == "naturalinc", "Natural Increase"),
         variable = replace(as.character(variable), variable == "internationalmig", "International Migration"),
         variable = replace(as.character(variable), variable == "domesticmig", "Domestic Migration"),
         variable = replace(as.character(variable), variable == "netmig", "Net Migration"),
         variable = replace(as.character(variable), variable == "npopchg", "Total Change")
         )

ggplot(ohio_totals, aes(x = year, y = value, color = variable)) +
  geom_line(size = 1.25) +
  ggtitle("Sources of Population Change") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title = element_blank(),
        legend.title = element_blank(),
        panel.grid.major.y = element_line(color = "black")
        ) +
  scale_y_continuous(label = comma)
```






# **Introduction to Prosperity Growth Rates**
# **+ Measures of Correlation**
Now that we know a little bit about the prosperity measurement and where certain states land, let's take a look at growth 
rates and a possible correlation between net population change.

 
```{r Setup}
library(dplyr)
library(RColorBrewer) #for fun colors!
library(ggplot2)
library(tidyr)
library(reshape2) #for melting, in case it's necessary.
library(lubridate) #for working with dates. 
library(esquisse) #for making some graphs with Shiny.
library(knitr)
library(kableExtra)
library(corrplot) #for help with correlations. 
library(Hmisc) #for specific correlation functions. 
library(corrr) #for more correlation functions. 
library(magrittr) 
library(plotly) #for fun graphs.
```


```{r}

prosperity<-prosperity %>%
  dplyr::rename(area_name = ï..area_name) %>%
  dplyr::select(area_name, area_group,score_2010,score_2011,score_2012,score_2013,score_2014,score_2015,score_2016,score_2017,score_2018,score_2019)
census2<-census%>%
  dplyr::select(-SUMLEV,-REGION,-DIVISION)
pillars<-pillars %>%
  dplyr::rename(area_name = ï..area_name) %>%
  dplyr::select(area_name, area_group,pillar_name,score_2010,score_2011,score_2012,score_2013,score_2014,score_2015,score_2016,score_2017,score_2018,score_2019)
pillars<- pillars %>%
  rename("2010" = score_2010,"2011" = score_2011, "2012" = score_2012, "2013" = score_2013, "2014" = score_2014, "2015" = score_2015, "2016" = score_2016, "2017" = score_2017, "2018" = score_2018, "2019" = score_2019)
pillar_time<-pillars %>%
  gather(Year, Value, "2010":"2019") %>%
  spread(pillar_name, Value)
regional<- pillars%>%
  filter(area_group=="Midwest") %>%
  gather(Year, Value, "2010":"2019") 
borderstates<- pillars%>%
  filter(area_name == "Ohio" | area_name == "Kentucky" | area_name == "Indiana" | area_name == "West Virginia" |area_name == "Pennsylvania" |
           area_name=="Michigan") %>%
  gather(Year, Value, "2010":"2019")
ohio<-pillars %>%
  filter(area_name=="Ohio") %>%
  gather(Year, Value, "2010":"2019")
```

```{r}
wanted_var<- c("Year","Business Environment","Economic Quality", "Education", "Governance", "Health", "Living Conditions", "Market Access & Infrastructure","Natural Environment", "Personal Freedom", "Safety & Security", "Social Capital")
pillar_names<-c("Business Environment","Economic Quality", "Education", "Governance", "Health", "Living Conditions", "Market Access & Infrastructure","Natural Environment", "Personal Freedom", "Safety & Security", "Social Capital")
border_states <- c("Ohio","Kentucky","Indiana","West Virginia","Pennsylvania","Michigan")
```

```{r}
pillar_census<-pillar_time%>%
  inner_join(census, by = c("area_name" = "NAME"), name= "State") %>%
  inner_join(prosperity, by = "area_name") %>%
  rename("2010avg"="score_2010","2011avg"="score_2011","2012avg"="score_2012","2013avg"="score_2013","2014avg"="score_2014","2015avg"="score_2015","2016avg"="score_2016","2017avg"="score_2017","2018avg"="score_2018","2019avg"="score_2019")
```

## Check the Correlations
One of our primary goals was to understand how population change was correlated with the particular measures of prosperity. 
To get there, we produced a correlation matrix, a correlation plot, and a simple bar chart. 
```{r}
corrs<- pillar_census %>%
  dplyr::select(NPOPCHG_2018,`Business Environment`,`Economic Quality`,Education,Governance,Health,`Living Conditions`,
         `Market Access & Infrastructure`,`Natural Environment`,`Personal Freedom`,`Safety & Security`,`Social Capital`) 
corrsc<- corrs %>%
  correlate() %>%
  focus(NPOPCHG_2018) 
corrsc
#corvarsp<-rcorr(cor(corrs))
#corvarsp
corrplot(cor(corrs), type="upper", order="hclust")
corrsg<- corrsc %>% 
  mutate(rowname = factor(rowname, levels = rowname[order(NPOPCHG_2018)])) %>%  
  ggplot(aes(x = rowname, y = NPOPCHG_2018, fill=rowname)) +
  geom_bar(stat = "identity") +
  ylab("Correlation with Population Change: 2018") +
  xlab("Prosperity Measure")+
  labs(title = "Correlation with Population Change", fill = "Prosperity Measure")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
corrsg
```

We can see that individual measures of prosperity have a relationship with population change, but not a big one. 
Half of the measures relate positively; half relate negatively.
Social capital and economic quality have the strongest correlations. 

## Fit a Regression
We also wanted to understand how we could best predict population change by prosperity.
To do this, we ran a normal linear regression.
```{r}
predpop<-lm(NPOPCHG_2018 ~ `Business Environment`+`Economic Quality`+Education+Governance+Health+
                `Living Conditions`+`Market Access & Infrastructure`+`Natural Environment`+
                `Personal Freedom`+`Safety & Security`+`Social Capital`,data=pillar_census)
summary(predpop)
```

In predicting population change with individual measures of prosperity, the model explains 24% of population change. 

## Examine the Growth Rate
Our final goal for this section was to see how the growth rate of Ohio fared against that of other states. 
Here, we create two new data frames to compare growth rates and we build two new models which incorporate growth.
```{r}
pillar_growth <- pillar_census %>%
  mutate(growth = (`2019avg` - `2011avg`)/`2011avg`) 
pg_ohio<-pillar_growth %>%
  filter(area_name %in% border_states) %>%
  group_by(area_name) %>%
  summarise(mean(growth))
pg_ohio
predpop3<- lm(NPOPCHG_2018 ~ growth, data=pillar_growth)
summary(predpop3)
predpop4<- lm (NPOPCHG_2018 ~ `Business Environment`+`Economic Quality`+Education+Governance+Health+
                `Living Conditions`+`Market Access & Infrastructure`+`Natural Environment`+
                `Personal Freedom`+`Safety & Security`+`Social Capital` + growth,data=pillar_growth)
summary(predpop4)
```

Among bordering states, Pennsylvania and Ohio have the lowest rates of prosperity growth.
Indiana has the highest. Putting growth into the same model as before explains another 5%. 

```{r}
pillar_time <- pillar_time %>%
  mutate(avgprosp = rowMeans(dplyr::select(.,4:14)))
pt_growth <- pillar_time %>%
  group_by(area_name) %>%
  mutate(growdiff = avgprosp - lag(avgprosp)) %>%
  mutate(perchange = growdiff / lag(avgprosp))
borders2<- borderstates %>%
  group_by(area_name,Year) %>%
  summarise(avg_prosp= mean(Value))
borders3<- pt_growth %>%
  filter(area_name %in% border_states)
regional2<- regional %>%
  group_by(area_name,Year) %>%
  summarise(avg_prosp= mean(Value))
ohio2<- pillar_census %>%
  filter(area_name =="Ohio")%>%
  select(-c(BIRTHS2010:RNETMIG2018))
ohio2<-ohio2[1:9,]
PopChange<-data.frame()
for (i in 27:35){
  PopChange[i,1]<-ohio2[1,i]
}
PopChange<-PopChange[27:35,1]
ohio5<-cbind(ohio2,"PopChange"=PopChange)
AverageScore<-data.frame()
for (i in 37:45){
  AverageScore[i,1]<-ohio5[1,i]
}
AverageScore<-AverageScore[37:45,1]
ohio6<-cbind(ohio5,"AverageScore"=AverageScore)
ohio7<-ohio6 %>%
  select(-c(CENSUS2010POP:`2019avg`)) %>%
  mutate(growdiff = AverageScore - lag(AverageScore)) %>%
  mutate(perchange = growdiff / lag(AverageScore))%>%
  mutate(popdiff = PopChange-lag(PopChange)) %>%
  mutate(perpop = popdiff / lag(PopChange))

rightside <- list(
  tickfont = list(color = "navy"),
  overlaying = "y",
  side = "right",
  title = "Average Score")
```

# Now, let's plot.

## Is prosperity in Ohio comparable to bordering states? 

```{r}
borders2 %>%
  group_by(area_name) %>%
  plot_ly(x= ~Year, y = ~avg_prosp, type = "scatter", mode= "lines+markers", split=~area_name, color=~area_name,
          colors="Paired") %>%
  layout(title = "Prosperity in Border States Over Time",
         xaxis=list(title="Year"),
         yaxis=list(title="Average Prosperity"))

```

## How does prosperity change in border states?  

```{r}
borders3 %>%
  group_by(area_name) %>%
  plot_ly(x= ~Year, y = ~perchange, type = "scatter", mode= "lines+markers", split=~area_name, color=~area_name,
          colors="Paired") %>%
  layout(title = "Percent Change in Prosperity Over Time",
         xaxis=list(title="Year"),
         yaxis=list(title="Average Prosperity"))
```

## What does prosperity look like in Ohio?

```{r}
ohio %>%
  plot_ly(x=~Year, y=~Value, color=~pillar_name, split=~pillar_name,
          colors="Paired",type="scatter",mode="lines+markers")%>%
  layout(title = "Prosperity in Ohio",
         xaxis=list(title="Year"),
         yaxis=list(title="Prosperity Measure"))
```

## Does the prosperity score have an effect on net population change? 

```{r}
ohio7 %>%
  plot_ly() %>%
  add_trace(x=~Year,y=~PopChange,name='Net Population Change',type="scatter",mode="lines+markers",color="blue") %>%
  add_trace(x=~Year,y=~AverageScore, name='Average Score',type="scatter",mode="lines+markers",color="green",yaxis="y2")%>%
  layout(title="Score + Population",
         xaxis=list(title="Year"),
         yaxis2=rightside)
```

## How do the percentage change in population and the average prosperity relate? 

```{r}
ohio7 %>%
  plot_ly() %>%
  add_trace(x=~Year,y=~perpop,name='Population % Change',type= "scatter",mode="lines+markers",color="blue") %>%
  add_trace(x=~Year,y=~AverageScore, name='Average Score',type="scatter",mode="lines+markers",color="green",yaxis="y2")%>%
  layout(title="Score + Population",
         xaxis=list(title="Year"),
         yaxis2=rightside)
```









```{R, echo = FALSE, message = FALSE}

library("leaflet")
library("htmlwidgets")
library("rgdal")
library("tigris")
library("tidyverse")
library("data.table")
library("reshape2")
library(dplyr)
library(ggplot2)
library(tidyr)
library(data.table)



```


# Part 1: Comparing Population Change

### Let's order our states by change in population from 2010 - 2028 

```{R}

census <- mutate(census, slope = (POPESTIMATE2018 - POPESTIMATE2010)/POPESTIMATE2010)

#Filter out for states 
census1 <- filter(census, SUMLEV == 40)

#Filter out unneeded columns
census2 <- select(census1, NAME, slope, starts_with("popestimate"))

#Arrange the data by slope
census3 <- arrange(census2, slope)

print(census3[8:18,1:2])
```

## What states had similar population growth to Ohio?

``` {r, echo = FALSE, message = FALSE}

library(maps)
library(mapdata)
```

``` {r}
map('state', fill = TRUE, col = "darkgray", border = "white", lwd = 1)
map(database = "usa", lwd = 1, add = TRUE)

map("state", "ohio", col = "blue",
    lwd = 1, fill = TRUE, add = TRUE)
map("state", c("new jersey", "new mexico", "michigan", "kansas", "pennsylvania"), col = "red",
    lwd = 1, fill = TRUE, add = TRUE)

title("Ohio Against the World")
```

## Were there any irregularities in the annual year-to-year changes in population that need to be understood?

```{R}
#To rearrange the data for graphing over time
census4 <- select(census3, NAME, POPESTIMATE2010:POPESTIMATE2018) %>%
  setnames(c("POPESTIMATE2010", "POPESTIMATE2011", "POPESTIMATE2012",
             "POPESTIMATE2013", "POPESTIMATE2014", "POPESTIMATE2015",
             "POPESTIMATE2016", "POPESTIMATE2017", "POPESTIMATE2018"), 
           c("Y2010", "Y2011", "Y2012", "Y2013",
             "Y2014", "Y2015", "Y2016", "Y2017", "Y2018")) %>%
  gather(key = "Year", value = "Population", Y2010:Y2018)

#Filter out to the states with similar growth to Ohio
similar_states <- filter(census4, NAME %in% c("Michigan", "New Jersey", "New Mexico", "Kansas", "Ohio", "	Pennsylvania"))

#Graphing the annual population changes from 2010 to 2018
ggplot(similar_states, mapping = aes(x = Year, y = Population, group = NAME, col= NAME)) + geom_line(size = 1) + geom_point(size = 2.5, shape = 17) 
```

## Not very interesting! 

### Although, compared to states that have had similar growth rates, Ohio has the largest population.

```{r, results='asis', echo=FALSE}
cat("\\newpage")
```


# Part 2: Comparing PI Change for Similar States

## How do our growth competitors compare by Prosperity Index?

```` {r}


#Clean up the PI data

PI_clean <- select(PI, area_name, score_2010:score_2018) %>%
  setnames(c("score_2010", "score_2011", "score_2012", "score_2013", "score_2014", 
              "score_2015", "score_2016", "score_2017","score_2018"),
          c("Y2010", "Y2011", "Y2012", "Y2013", "Y2014", "Y2015", "Y2016", "Y2017", "Y2018")) %>%
  gather(key = "Year", value = "PI_score", Y2010:Y2018)

#Filter out the states that are not similar to Ohio's population growth from 2010 to 2018
PI_sim_states <- filter(PI_clean, area_name %in% c("Michigan", "New Jersey", "New Mexico", "Kansas", 
                                                         "Ohio", "	Pennsylvania"))

#Graph PI Changes for Similar States
ggplot(PI_sim_states, mapping = aes(x = Year, y = PI_score, group = area_name, col = area_name)) + geom_line(size = 1) + geom_point(size = 2.5, shape = 17) 

```

## Any similarities? Differences?

## Ohio has the largest population and consistently the second lowest PI score, but we have the second largest increase in PI from 2010-2018.


``` {r}

#To calculate the changes in PI for each similar state from 2010-2018

PI_slope <- mutate(PI, slopePI = (score_2018 - score_2010)/score_2010) %>% setnames("area_name", "State")

PI_slope1 <- filter(PI_slope, State %in% c("Michigan", "New Jersey", "New Mexico", "Kansas", 
                                    "Ohio", "	Pennsylvania"))
PI_slope2 <- select(PI_slope1, State, slopePI, score_2010:score_2018)

PI_slope3 <- arrange(PI_slope2, slopePI)

print(PI_slope3[1:5,1:2])

```
## Whose PI was the highest?
 
 
``` {r, echo = FALSE, message = FALSE}
land_area <- read_csv("Land_Area_Data.CSV") 

```

# Part 3: Density

``` {r, message = FALSE, warning = FALSE}

#Creat new data frame for testing relationship between density and changes in PI

joined_data <- PI_slope %>% inner_join(land_area) 

ggplot(data = joined_data, aes(x = slopePI, y = Density)) + 
  geom_point(size = 2.5, shape = 17) + 
  geom_smooth(method = lm, se = FALSE, size = 1)

```

## How is that outlier skewing the data?

``` {r, message = FALSE}
#Delete Washington DC because it was skewing data
joined_cleaner <- joined_data[-c(9),]



#graphing relationship between change in PI and density
ggplot(data = joined_cleaner, aes(x = slopePI, y = Density)) + 
  geom_point(size = 2.5, shape = 17) + 
  geom_smooth(method = lm, se = FALSE, size = 1)

#is there significance?
Regression_test <- lm(slopePI ~ Density + Land.Area, data = joined_cleaner)
summary(Regression_test)

```

## Density is not really significant nor is the graph extremely interesting.

## What about Land Area?

``` {r}
ggplot(data = joined_cleaner, aes(x = slopePI, y = Land.Area)) + 
  geom_point(size = 2.5, shape = 17) + 
  geom_smooth(method = lm, se = FALSE, size = 1)

```

## Let's remove the Land Area outlier to see what happens.

``` {r}
joined_cleaner2 <- joined_cleaner[-c(2),]

ggplot(data = joined_cleaner2, aes(x = slopePI, y = Land.Area)) + 
  geom_point(size = 2.5, shape = 17) + geom_smooth(method = lm, se = FALSE)
  

Regression_test2 <- lm(slopePI ~ Density + Land.Area, data = joined_cleaner2)
summary(Regression_test2)

```

## Land Area becomes a less significant variable.









## Comparing Ohio to Indiana and Kentucky

``` {r, error = FALSE}
#non interactive radar plot
#install.packages("hrbrthemes")
#install.packages("fmsb")
#install.packages("colormap")

library(tidyverse)
library(data.table) #setnames
library(viridis) #this does something with colors
library(hrbrthemes) #this does something with colors
library(fmsb) #has the radar plot
library(colormap) #more colors

state <- read.csv("Pillars_x11.csv")

#tidying the data
state.radar <- select(state, area_code, pillar_name, score_2019) %>%
  filter(area_code %in% c("OH", "IN", "KY")) %>%
  spread(pillar_name, score_2019) %>%
  setnames("area_code", "PI") %>%
  select("Business Environment":"Social Capital")
row.names(state.radar) <- c("IN", "KY", "OH")

# To use the fmsb package, I have to add 2 lines to the dataframe: the max and min of each topic to show on the plot!
state.radar2 <-rbind(rep(70,11) , rep(0,11) , state.radar)


# Prepare color
colors_border=c(rgb(0.8,0.2,0.5,0.9), rgb(0.5,0.8,0.2,0.9), rgb(0.2,0.5,0.5,0.9))
colors_in=c( rgb(0.8,0.2,0.5,0.4), rgb(0.5,0.8,0.2,0.4), rgb(0.5,0.5,0.5,0.4))

#Prepare names
statenames <- c("Indiana", "Kentucky", "Ohio")

# Custom the radarChart !


radarplot <- radarchart(state.radar2, axistype = 1, centerzero = TRUE, seg = 7,
            #custom polygon
            pcol = colors_border, pfcol = colors_in, plwd=4, plty=1,
            #custom the grid
            cglcol="grey", cglty=1, axislabcol="grey", caxislabels=seq(0,70,10), cglwd=1.1,
            #custom labels
            vlcex=0.8, title = "Prosperity Index State Comparison 2018") 
            # legend
            #legend(1, 1, legend = statenames, bty = "n", pch=20 , col=colors_border, cex=0.9, pt.cex=1.6)

```

### Comparing Prosperity Index Scores for 2018
```{r}
#plot(radarplot)
```

## Changes in PI and Death Rate
If a state becomes safer and healthier (as represented by the PI scores: Health, Living Conditions, and Safety and Security), does the death rate decline?

```{r}
library(dplyr) 
library(ggplot2)
library(tidyr)
library(data.table) #to get setnames function



#indexing the death rate from 2011, setting it to percent change so it is on the 
#same scale as the prosperity index, tidying up the data
state.death <- select(state.data, NAME, RDEATH2011:RDEATH2018) %>%
  setnames(c("RDEATH2011", "RDEATH2012", "RDEATH2013",
             "RDEATH2014", "RDEATH2015", "RDEATH2016",
             "RDEATH2017", "RDEATH2018"), 
           c("Y2011", "Y2012", "Y2013", "Y2014",
             "Y2015", "Y2016", "Y2017", "Y2018")) %>%
  mutate( change.2012 = (Y2012 - Y2011) / Y2011 * 100,
          change.2013 = (Y2013 - Y2011) / Y2011 * 100,
          change.2014 = (Y2014 - Y2011) / Y2011 * 100,
          change.2015 = (Y2015 - Y2011) / Y2011 * 100,
          change.2016 = (Y2016 - Y2011) / Y2011 * 100,
          change.2017 = (Y2017 - Y2011) / Y2011 * 100,
          change.2018 = (Y2018 - Y2011) / Y2011 * 100)

#reorganizing the data so it is easier to plot
state.death.change <- select(state.death, NAME, change.2012:change.2018) %>%
  gather(key = "year.change", value = "Change in Death Rate",
         "change.2012":"change.2018")

#Indexing prosperity index data and tidying
prosperity.index.filter <- select(state, area_name, pillar_name, score_2011:score_2018) %>%
  filter(pillar_name %in% c("Health", "Living Conditions", 
                            "Safety & Security")) %>%
  mutate(change.2012 = (score_2012 - score_2011) / score_2011 * 100,
         change.2013 = (score_2013 - score_2011) / score_2011 * 100,
         change.2014 = (score_2014 - score_2011) / score_2011 * 100,
         change.2015 = (score_2015 - score_2011) / score_2011 * 100,
         change.2016 = (score_2016 - score_2011) / score_2011 * 100,
         change.2017 = (score_2017 - score_2011) / score_2011 * 100,
         change.2018 = (score_2018 - score_2011) / score_2011 * 100)

prosperity.death <- select(prosperity.index.filter, area_name, pillar_name, 
                           change.2012:change.2018) %>%
  gather(key = "year.change", value = "percent.change.pi", 
         "change.2012":"change.2018") %>%
  spread(pillar_name, percent.change.pi)

#joining to to dataframes
new.df <- inner_join(prosperity.death, state.death.change, 
                     by = c("area_name" = "NAME", "year.change")) %>%
  gather(key = "category", value = "percent.change", "Health":
           "Change in Death Rate")

#filtering to just what I want
new.df2 <- filter(new.df, area_name %in% c("Ohio", "Kentucky", "Indiana")) 

#adding a facet wrap by category
chart1 <- ggplot(new.df2, aes(x = year.change, y = percent.change, group = interaction(area_name, category), color = area_name)) +
  geom_line(aes(linetype = category)) +
  geom_point(aes(shape = area_name)) +
  facet_wrap(~category)

finchart <- chart1 + scale_x_discrete(labels=c("change.2012" = "2012", "change.2013" = "2013",
                                   "change.2014" = "2014", "change.2015" = "2015",
                                   "change.2016" = "2016", "change.2017" = "2017",
                                   "change.2018" = "2018")) +
  xlab("Year to Year Change") +
  ylab("Percentage Change") +
  scale_color_discrete(name = "State") +
  scale_linetype_discrete(name = "Category") +
  scale_shape(guide = FALSE) +
  ggtitle("PI Comparisons by State and PI Category")

```

### Plot
Comparing PI Scores and Death Rates for Ohio, Indiana, and Kentucky from 2011-2018: 

```{r}
plot (finchart)

```

## Ohio's PI and Death Rate
Let's Examine Ohio more closely over time, let's organize our data first:

```{r}


#indexing the death rate from 2011, setting it to percent change so it is on the 
#same scale as the prosperity index, tidying up the data
state.death <- select(state.data, NAME, RDEATH2011:RDEATH2018) %>%
  setnames(c("RDEATH2011", "RDEATH2012", "RDEATH2013",
             "RDEATH2014", "RDEATH2015", "RDEATH2016",
             "RDEATH2017", "RDEATH2018"), 
           c("Y2011", "Y2012", "Y2013", "Y2014",
             "Y2015", "Y2016", "Y2017", "Y2018")) %>%
  mutate( change.2012 = (Y2012 - Y2011) / Y2011 * 100,
          change.2013 = (Y2013 - Y2011) / Y2011 * 100,
          change.2014 = (Y2014 - Y2011) / Y2011 * 100,
          change.2015 = (Y2015 - Y2011) / Y2011 * 100,
          change.2016 = (Y2016 - Y2011) / Y2011 * 100,
          change.2017 = (Y2017 - Y2011) / Y2011 * 100,
          change.2018 = (Y2018 - Y2011) / Y2011 * 100)

#reorganizing the data so it is easier to plot
state.death.change <- select(state.death, NAME, change.2012:change.2018) %>%
  gather(key = "year.change", value = "Change in Death Rate",
         "change.2012":"change.2018")

#Indexing prosperity index data and tidying
prosperity.index.filter <- select(state, area_name, pillar_name, score_2011:score_2018) %>%
  filter(pillar_name %in% c("Health", "Living Conditions", 
                            "Safety & Security")) %>%
  mutate(change.2012 = (score_2012 - score_2011) / score_2011 * 100,
         change.2013 = (score_2013 - score_2011) / score_2011 * 100,
         change.2014 = (score_2014 - score_2011) / score_2011 * 100,
         change.2015 = (score_2015 - score_2011) / score_2011 * 100,
         change.2016 = (score_2016 - score_2011) / score_2011 * 100,
         change.2017 = (score_2017 - score_2011) / score_2011 * 100,
         change.2018 = (score_2018 - score_2011) / score_2011 * 100)

prosperity.death <- select(prosperity.index.filter, area_name, pillar_name, 
                           change.2012:change.2018) %>%
  gather(key = "year.change", value = "percent.change.pi", 
         "change.2012":"change.2018") %>%
  spread(pillar_name, percent.change.pi)

#joining to to dataframes
new.df <- inner_join(prosperity.death, state.death.change, 
                     by = c("area_name" = "NAME", "year.change")) %>%
  gather(key = "category", value = "percent.change", "Health":
           "Change in Death Rate")

#filtering to just Ohio death and one category

new.df.oh.h <- filter(new.df, area_name == "Ohio") %>%
               filter(category %in% c("Health", "Change in Death Rate"))

new.df.oh.lc <- filter(new.df, area_name == "Ohio") %>%
  filter(category %in% c("Living Conditions", "Change in Death Rate"))

new.df.oh.ss <- filter(new.df, area_name == "Ohio") %>%
  filter(category %in% c("Safety & Security", "Change in Death Rate"))

```

### Ohio Charts
Here we can examine how the Health PI score and death rate have changed over time.
Let's start with Health

```{r}
#ohio health
chart.ohio.h <- ggplot(new.df.oh.h, aes(x = year.change, y = percent.change, group = category, color = category)) +
  geom_line()


ohiochart1 <- chart.ohio.h + scale_x_discrete(labels=c("change.2012" = "2012", "change.2013" = "2013",
                                   "change.2014" = "2014", "change.2015" = "2015",
                                   "change.2016" = "2016", "change.2017" = "2017",
                                   "change.2018" = "2018")) +
  xlab("Year to Year Change") +
  ylab("Percentage Change") +
  scale_color_discrete(name = "Category") +
  ggtitle("Ohio Comparison of Changes in Health Ranking and Death Rate")

```

### Plot

```{r}
plot(ohiochart1)
```

Now let's examine how Living Conditions have changed

```{r}
#ohio living conditions 

chart.ohio.lc <- ggplot(new.df.oh.lc, aes(x = year.change, y = percent.change, group = category, color = category)) +
  geom_line()


ohiochart2 <- chart.ohio.lc + scale_x_discrete(labels=c("change.2012" = "2012", "change.2013" = "2013",
                                         "change.2014" = "2014", "change.2015" = "2015",
                                         "change.2016" = "2016", "change.2017" = "2017",
                                         "change.2018" = "2018")) +
  xlab("Year to Year Change") +
  ylab("Percentage Change") +
  scale_color_discrete(name = "Category") +
  ggtitle("Ohio Comparison of Changes in Living Conditions Rating and Death Rate")
```

### Plot

```{r}
plot(ohiochart2)
```

And finally Safety and Security

```{r}
#ohio safety and security 
chart.ohio.ss <- ggplot(new.df.oh.ss, aes(x = year.change, y = percent.change, group = category, color = category)) +
  geom_line()


ohiochart3 <- chart.ohio.ss + scale_x_discrete(labels=c("change.2012" = "2012", "change.2013" = "2013",
                                          "change.2014" = "2014", "change.2015" = "2015",
                                          "change.2016" = "2016", "change.2017" = "2017",
                                          "change.2018" = "2018")) +
  xlab("Year to Year Change") +
  ylab("Percentage Change") +
  scale_color_discrete(name = "Category") +
  ggtitle("Ohio Comparison of Changes in Safety & Security Rating and Death Rate")
```

### Plot

```{r}
plot(ohiochart3)
```

<!-- ```{r} -->
<!-- prosperity.index.filter <- state %>% -->
<!--   select(area_name, pillar_name, score_2009:score_2018) %>% -->
<!--   filter(area_name == "Ohio") %>% -->
<!--   gather(key = "year", value = "year.score", "score_2009":"score_2018") -->

<!-- ohiochart <- ggplot(prosperity.index.filter, aes(x = year, y = year.score, group = pillar_name, color = pillar_name)) + -->
<!--   geom_line() -->

<!-- ohiochart + scale_x_discrete(labels=c("score_2009" = "2009", "score_2010" = "2010",  -->
<!--                                       "score_2011" = "2011", "score_2012" = "2012", -->
<!--                                       "score_2013" = "2013", "score_2014" = "2014",  -->
<!--                                       "score_2015" = "2015", "score_2016" = "2016", -->
<!--                                       "score_2017" = "2017", "score_2018" = "2018")) + -->
<!--   xlab("Year to Year Change") + -->
<!--   ylab("Prosperity Index Score") + -->
<!--   scale_color_discrete(name = "Pillar") + -->
<!--   ggtitle("Ohio PI Scores Change") -->
<!-- ``` -->

```{r}
sessionInfo()
```

